---
title: "CRC_point_process"
author: "Charlotte M. Jones-Todd"
date: "September 2017"
output: md_document
bibliography: ~/MEGA/my_styles/bib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE)
## devtools::install_github("b-steve/palm") ## version not on CRAN exports parents
## rmarkdown::render('CRC_point_process.Rmd', output_file = '~/Git/examples/CRC_point_process.md') 
```

# Simulated examples

Parameter estimation for the Thomas, Matérn, and void process is carried out by the R package **palm** (@stevenson2017palm), which is available on CRAN. All models are fitted using either the **fit.ns()** function, for NSPP, or the **fit.void()** function, for the void process. 

Below, code examples demonstrate using functionality of the **palm** (@stevenson2017palm) package to simulate from and fit a two-dimensional point process. This is done for of each of the types mentioned above. The simulated data for each process are shown---parents are plotted at grey crosses and daughters by black dots---along with the fitted (solid lines) and empirical (dashed lines) Palm intensity functions. 

Please note that parameters of the Palm intensity functions may differ, in name only, from the descriptions in @charlotte2017cancer.

```{r library,eval = TRUE,message=FALSE}
library(palm)
```

```{r functions,eval = TRUE, echo = FALSE}
## A simple function that plots simulated data showing parent locations
## its only argument is the simulated list object which must contain named
## elements *points* and *parents*.
plot.palm.sim <- function(x){
  plot(x$points,pch = 20,xlab = "",ylab = "",xaxt = "n",yaxt = "n",asp = 1)
  points(x$parents, pch = 4, col = "grey")
}
```

## Neyman-Scott point processes

Cluster processes where unobserved parent points randomly generate daughters centered at their unobserved location. The spatial distribution of daughters may eiher follow a bivariate normal distribution (Thomas) or be uniformally distributed in a circle (Mat\'ern).

### Thomas process

The Palm intensity function of a Thomas process may be characterised by the parameter vector $\boldsymbol{\theta} = (D,\lambda,\sigma)$. Here $D$ is the density of parents, $\lambda$ gives the expected number of daughters per parent, and $\sigma$ is the standard deviation of the bivariate normal distribution by which daughters are scattered around their parents.

```{r Thomas}
set.seed(1234)
lims <- rbind(c(0, 1),c(0,1)) ## 2D limits of domain (i.e., unit square)
thomas <- sim.ns(c(D = 7, lambda = 8, sigma = 0.05), lims = lims) ## simulate a Thomas process
fit.thomas <- fit.ns(thomas$points,lims = lims, R = 0.5) ## fit a Thomas process
coef(fit.thomas)
```

```{r plot thomas,echo = FALSE,fig.height=3}
layout(matrix(c(1,2,2),nrow = 1))
plot.palm.sim(thomas)
plot(fit.thomas)
```

### Matérn process

The Palm intensity function of a Matérn process may be characterised by the parameter vector $\boldsymbol{\theta} = (D,\lambda,\tau)$. Here $D$ is the density of parents, $\lambda$ gives the expected number of daughters per parent, and $\tau$ is the radius of the sphere, centered at a parent location, within which daughters are uniformally scattered.


```{r Matern}
set.seed(2344)
lims <- rbind(c(0, 1),c(0,1)) ## 2D limits of domain (i.e., unit square)
matern <- sim.ns(c(D = 7, lambda = 8, tau = 0.05), lims = lims,disp = "uniform") ## simulate a Matern process
fit.matern <- fit.ns(matern$points,lims = lims, R = 0.5, disp = "uniform") ## fit a Matern process
coef(fit.matern)
```

```{r plot matern,echo = FALSE,fig.height=3}
layout(matrix(c(1,2,2),nrow = 1))
plot.palm.sim(matern)
plot(fit.matern)
```

## Void point process

The Palm intensity function of a void process may be characterised by the parameter vector $\boldsymbol{\theta} = (D_c,D_p,\tau)$. Here $D_c$ is the density of children, $D_p$ is the density of parents, and $\tau$ is the radius of the voids centered at each parent. 

```{r void}
set.seed(3454)
lims <- rbind(c(0, 1),c(0,1)) ## 2D limits of domain (i.e., unit square)
void <- sim.void(pars = c(Dc = 300, Dp = 10,tau = 0.075),lims = lims) ## simulate a void process
fit.void <- fit.void(points = void$points, lims = lims, R = 0.5,
                     bounds = list(Dc = c(280,320), Dp = c(8,12), tau = c(0,0.2))) ## fit a void process
coef(fit.void)
```

```{r plot void,echo = FALSE,fig.height=3}
layout(matrix(c(1,2,2),nrow = 1))
plot.palm.sim(void)
plot(fit.void)
```

## Variance estimation

Variance estimation is achieved via a parametric bootstrap, which can be carried out through the use of the **boot.palm()** function. For example, running **boot.palm(fit.thomas,N = 1000)** will perform 1000 bootstrap resamples of the fitted thomas process in order to estimate standard errors of the parameters.

```{r testing,eval = FALSE,echo = FALSE}
library(gapski)
points<-sim.gap(pars=c(D=10,lambda=300,R=0.075))
fit.gap<-fit.gap(points = points,lims = rbind(c(0,1),c(0,1)),trunc = 0.5,D.sv = 10,D.bounds=c(8,12))
library(nspp)
points <- sim.ns(pars=c(D=7, lims = rbind(c(0, 1), c(0, 1)), child.disp=0.05,child.par=8),
                 dispersion="uniform",rchild=rpois)
child.dist = list(mean = function(x) x, var = function(x) x,sv = 8, bounds = c(1e-6, nrow(points)))
fit.mat.pois.2D <- fit.ns(points = points, lims = rbind(c(0, 1), c(0, 1)),R = 0.5, dispersion="uniform",child.dist = child.dist)
```

# CRC data

As per @charlotte2017cancer below is an digital image of a tissue section from a CRC patient (left hand plot of the Figure), the tumour and stroma structures of the tissue sections are coloured in red and green respectively. The far right plot of shows the point pattern formed by the tumour and stroma cell nuclei, black and grey respectively, of the same tissue section.

![Illustration of the image analysis of one patient's slide which enables the pinpointing of nuclei. Left: Composite immunofluorescence digital image showing Tumour (red), Stroma (green) and all nuclei (blue). Middle: Image analysis mask overlay from automatic machine learnt segmentation of the digital image. Tumour (purple), stroma (turquoise), necrosis (yellow). Right: Point pattern formed by the nuclei of the tumour (black) and stroma (grey) cells shown in the previous two images.](CRC_point_process_files/figure-markdown_strict/cancer.png)
```{r load,eval = FALSE,echo = FALSE}
load("~/mega/papers/cancer/cancerdata.RData")

```

```{r data setup,echo = TRUE,eval = FALSE}
library(palm)
## to allow parallel computing from R
library(foreach)
library(doSNOW)
R <- 0.5
n <- length(Tu)
ncores <- 5

## tumour locs
points<-list()
## stroma locs
pointss<-list()
for(i in 1:n){
    points[[i]] <- cbind(Tu[[i]]$inner_x,Tu[[i]]$inner_y)/2500
    pointss[[i]] <- cbind(St[[i]]$inner_x,St[[i]]$inner_y)/2500
}


## Set up results file
## ID, grades, and mortality index
ID <- grade <- mort <- rep(NA,n)

for(i in 1:n){
    grade[i]<-as.character(Tu[[i]]$Grade[1])
    mort[i]<-Tu[[i]]$Mortality[1]
    ID[i]<-Tu[[i]][1,1]
}

resultDF <- data.frame(ID = ID,grade = grade, mort = mort)

## set up results columns
resultDF$thomas.T.D <- resultDF$thomas.T.lam <- resultDF$thomas.T.sig <- rep(NA,n)
resultDF$thomas.S.D <- resultDF$thomas.S.lam <- resultDF$thomas.S.sig <- rep(NA,n)
resultDF$matern.T.D <- resultDF$matern.T.lam <- resultDF$matern.T.tau <- rep(NA,n)
resultDF$matern.S.D <- resultDF$matern.S.lam <- resultDF$matern.S.tau <- rep(NA,n)
resultDF$void.T.Dp <- resultDF$void.T.Dc <- resultDF$void.T.tau <- rep(NA,n)
resultDF$void.S.Dp <- resultDF$void.S.Dc <- resultDF$void.S.tau <- rep(NA,n)
 
## write patient info out to a file
## write.csv(resultDF,file = "resultsDF.csv",row.names = FALSE)

## change number of cores used
## create a cluster with ncores cores
cl <- makeSOCKcluster(ncores)
registerDoSNOW(cl)
progress <-  function(x) cat(sprintf("T Thomas model %d \n", x))
opts <- list(progress=progress)
     
```

## Fit Thomas process

```{r fit Thomas CRC,eval = FALSE}
### Tumour

### Tumour

resultDF[,c(6,5,4)] <- foreach(i = 1:n, 
              .combine = "rbind", 
              .packages = "palm",
              .options.snow=opts) %dopar% {
    fit <- fit.ns(points=points[[i]],lims=rbind(c(0,1),c(0,1)),R = R,
                  start = c(sigma = start[i,4], lambda = start[i,5], D = start[i,6]))
    coef(fit)[1:3]
}
### Stoma
progress <-  function(x) cat(sprintf("S Thomas model %d \n", x))
opts <- list(progress=progress)

resultDF[,c(9,8,7)] <- foreach(i = 1:n, 
              .combine = "rbind", 
              .packages = "palm",
              .options.snow=opts) %dopar% {
    fit <- fit.ns(points=pointss[[i]],lims=rbind(c(0,1),c(0,1)),R = R,
                  start = c(sigma = start[i,7], lambda = start[i,8], D = start[i,9]))
    coef(fit)[1:3]
}
## write patient info out to a file
## write.csv(resultDF,file = "resultsDF.csv",row.names = FALSE)

```


## Fit Matérn process

```{r fit Matern CRC,eval = FALSE}
### Tumour
progress <-  function(x) cat(sprintf("T Matern model %d \n", x))
opts <- list(progress=progress)

resultDF[,c(12,11,10)] <- foreach(i = 1:n, 
              .combine = "rbind", 
              .packages = "palm",
              .options.snow=opts) %dopar% {
    fit <- fit.ns(points=points[[i]],lims=rbind(c(0,1),c(0,1)),R = R,disp = "uniform",
                  start = c(tau = start[i,10], lambda = start[i,11], D = start[i,12]))
    coef(fit)[1:3]
}


### Stoma

progress <-  function(x) cat(sprintf("S Matern model %d \n", x))
opts <- list(progress=progress)

resultDF[,c(15,14,13)] <- foreach(i = 1:n, 
              .combine = "rbind", 
              .packages = "palm",
              .options.snow=opts) %dopar% {
                  fit <- fit.ns(points=pointss[[i]],lims=rbind(c(0,1),c(0,1)),R = R,disp = "uniform",
                                start = c(tau = start[i,13], lambda = start[i,14], D = start[i,15]))
    coef(fit)[1:3]
}
## write patient info out to a file
## write.csv(resultDF,file = "resultsDF.csv",row.names = FALSE)


```

## Fit Void process
```{r fit Void CRC,eval = FALSE}
### Tumour

progress <-  function(x) cat(sprintf("T void model %d \n", x))
opts <- list(progress=progress)

resultDF[,c(18,17,16)] <- foreach(i = 1:n, 
              .combine = "rbind", 
              .packages = "palm",
              .options.snow=opts) %dopar% {
                  fit <- fit.void(points=points[[i]],lims=rbind(c(0,1),c(0,1)),R = R,
                                  bounds = list(Dp = c(0,70)),
                                  start = c(tau = start[i,16], Dc = start[i,17], Dp = start[i,18]))
    coef(fit)[1:3]
}


### Stoma

progress <-  function(x) cat(sprintf("S void model %d \n", x))
opts <- list(progress=progress)

resultDF[,c(21,20,19)] <- foreach(i = 1:n, 
              .combine = "rbind", 
              .packages = "palm",
              .options.snow=opts) %dopar% {
                  fit <- fit.void(points=pointss[[i]],lims=rbind(c(0,1),c(0,1)),R = R,
                                  bounds = list(Dp = c(0,70)),
                                  start = c(tau = start[i,19], Dc = start[i,20], Dp = start[i,21]))
    coef(fit)[1:3]
}
## write patient info out to a file
## write.csv(resultDF,file = "resultsDF.csv",row.names = FALSE)
```

## Bootstrap functions

```{r bootstrap,eval=FALSE,echo=TRUE}
## read in csv of parameters (as above)
df <- read.csv("resultsDF.csv")


############# Heirarchical bootstrap
resample <- function(dat, cluster, replace) {
  # sample the clustering factor
  cls <- sample(unique(dat[[cluster[1]]]), replace=replace[1])
  # subset on the sampled clustering factors
  sub <- lapply(cls, function(b) subset(dat, dat[[cluster[1]]]==b))
  # join and return samples
  do.call(rbind, sub)

}

## Function to do a heirarchical bootstrap for CRC data.
## This function resamples based on a nested level (i.e., as each patient has multiple slides)
## Arguments are, 1) data a named list of numeric vectors to be bootstrapped 2) nsim times based on the factor
## vector 3) patient
## By default the "median" of the bootstrapped resamples (again median resamples) for each list element (data) are returned
## along with 95% quantiles

bootstrap <- function(data, nsim, patient, value = "median"){
    pb <- txtProgressBar(style = 3)
    cluster = "patient"
    boots <- list()
    for(i in 1:length(data)){
        dat <- data.frame(measurement = data[[i]],patient = patient)
        boots[[i]] <- replicate(nsim, median(resample(dat, cluster, TRUE)$measurement,na.rm = TRUE))
        setTxtProgressBar(pb,i)
    }
    close(pb)
    val <- sapply(boots,value)
    quantiles <- sapply(boots,quantile,c(0.025,0.975))
    res <- rbind(val = val, quantiles)
    colnames(res) <- names(data)
    return(res)
}

## Function to chose the data to split, and the level at which to split to perform the heirarchical bootstrap above.
## name should be a character vector of the data one wishes to bootstrap.
## by a single character of the level at which to split (i.e., mortality of patient).
## ID a single character specifying the name of the column in df whuch contains
## the patient IDs.
## df a matrix of data with columns named name, by and ID as above.
## nsim an integer specifying the number of bootstraps to do.
## the function returns a nested list of the output of the bootstrap function (above)
## for each chosen level.

fun <- function(name, by, ID, df, nsim){
    d <- list()
    for (i in 1:length(name)){
        d[[i]] <-  split(df[[name[i]]],df[[by]])
    }
    pat <- split(df[[ID]],df[[by]])
    names(d) <- name
    boots <- list()
    for (i in 1:length(d)){
        boots[[i]] <- list()
        for(j in 1:length(d[[i]])){
            boots[[i]][[j]] <- bootstrap(data = d[[i]][j],nsim = nsim, patient = pat[[j]])
        }
    }
    names(boots) <- name
    boots
}

```
For example to bootstrap the void process dispersion parameters for both the Tumour and Stroma nuclei, the following code should be run.
```{r example, echo=TRUE,eval=FALSE}
res <- fun(c("void.T.tau","void.S.tau"),"grade","ID", df,1000)

```
# References