---
title: "CRC_point_process"
author: "Charlotte M. Jones-Todd"
date: "September 2017"
output: md_document
bibliography: ~/mega/my_styles/bib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE)
## devtools::install_github("b-steve/palm") ## version not on CRAN exports parents
## rmarkdown::render('CRC_point_process.Rmd', output_file = '~/Git/examples/CRC_point_process.html') 
```

# Simulated examples

Parameter estimation for the Thomas, Matérn, and void process is carried out by the R package **palm**, which is available on CRAN. All models are fitted using either the **fit.ns()** function, for NSPP, or the **fit.void()** function, for the void process. 

Below, code examples demonstrate using functionality of the **palm** package to simulate from and fit a two-dimensional point process. This is done for of each of the types mentioned above. The simulated data for each process are shown---parents are plotted at grey crosses and daughters by black dots---along with the fitted (solid lines) and empirical (dashed lines) Palm intensity functions. 

Please note that parameters of the Palm intensity functions may differ, in name only, from the descriptions in @charlotte2017cancer.

```{r library,eval = TRUE,message=FALSE}
library(palm)
```

```{r functions,eval = TRUE, echo = FALSE}
## A simple function that plots simulated data showing parent locations
## its only argument is the simulated list object which must contain named
## elements *points* and *parents*.
plot.palm.sim <- function(x){
  plot(x$points,pch = 20,xlab = "",ylab = "",xaxt = "n",yaxt = "n",asp = 1)
  points(x$parents, pch = 4, col = "grey")
}
```

## Neyman-Scott point processes

Cluster processes where unobserved parent points randomly generate daughters centered at their unobserved location. The spatial distribution of daughters may eiher follow a bivariate normal distribution (Thomas) or be uniformally distributed in a circle (Mat\'ern).

### Thomas process

The Palm intensity function of a Thomas process may be characterised by the parameter vector $\boldsymbol{\theta} = (D,\lambda,\sigma)$. Here $D$ is the density of parents, $\lambda$ gives the expected number of daughters per parent, and $\sigma$ is the standard deviation of the bivariate normal distribution by which daughters are scattered around their parents.

```{r Thomas}
lims <- rbind(c(0, 1),c(0,1)) ## 2D limits of domain (i.e., unit square)
thomas <- sim.ns(c(D = 7, lambda = 8, sigma = 0.05), lims = lims) ## simulate a Thomas process
fit.thomas <- fit.ns(thomas$points,lims = lims, R = 0.5) ## fit a Thomas process
coef(fit.thomas)
```

```{r plot thomas,echo = FALSE,fig.height=3}
layout(matrix(c(1,2,2),nrow = 1))
plot.palm.sim(thomas)
plot(fit.thomas)
```

### Matérn process

The Palm intensity function of a Matérn process may be characterised by the parameter vector $\boldsymbol{\theta} = (D,\lambda,\tau)$. Here $D$ is the density of parents, $\lambda$ gives the expected number of daughters per parent, and $\tau$ is the radius of the sphere, centered at a parent location, within which daughters are uniformally scattered.


```{r Matern}
lims <- rbind(c(0, 1),c(0,1)) ## 2D limits of domain (i.e., unit square)
matern <- sim.ns(c(D = 7, lambda = 8, tau = 0.05), lims = lims,disp = "uniform") ## simulate a Matern process
fit.matern <- fit.ns(matern$points,lims = lims, R = 0.5, disp = "uniform") ## fit a Matern process
coef(fit.matern)
```

```{r plot matern,echo = FALSE,fig.height=3}
layout(matrix(c(1,2,2),nrow = 1))
plot.palm.sim(matern)
plot(fit.matern)
```

## Void point process

The Palm intensity function of a void process may be characterised by the parameter vector $\boldsymbol{\theta} = (D_c,D_p,\tau)$. Here $D_c$ is the density of children, $D_p$ is the density of parents, and $\tau$ is the radius of the voids centered at each parent. 

```{r void}
lims <- rbind(c(0, 1),c(0,1)) ## 2D limits of domain (i.e., unit square)
void <- sim.void(pars = c(Dc = 300, Dp = 10,tau = 0.075),lims = lims) ## simulate a void process
fit.void <- fit.void(points = void$points, lims = lims, R = 0.5,
                     bounds = list(Dc = c(280,320), Dp = c(8,12), tau = c(0,0.2))) ## fit a void process
coef(fit.void)
```

```{r plot void,echo = FALSE,fig.height=3}
layout(matrix(c(1,2,2),nrow = 1))
plot.palm.sim(void)
plot(fit.void)
```

## Variance estimation

Variance estimation is achieved via a parametric bootstrap, which can be carried out through the use of the **boot.palm()** function. For example, running **boot.palm(fit.thomas,N = 1000)** will perform 1000 bootstrap resamples of the fitted thomas process in order to estimate standard errors of the parameters.

```{r testing,eval = FALSE,echo = FALSE}
library(gapski)
points<-sim.gap(pars=c(D=10,lambda=300,R=0.075))
fit.gap<-fit.gap(points = points,lims = rbind(c(0,1),c(0,1)),trunc = 0.5,D.sv = 10,D.bounds=c(8,12))
library(nspp)
points <- sim.ns(pars=c(D=7, lims = rbind(c(0, 1), c(0, 1)), child.disp=0.05,child.par=8),
                 dispersion="uniform",rchild=rpois)
child.dist = list(mean = function(x) x, var = function(x) x,sv = 8, bounds = c(1e-6, nrow(points)))
fit.mat.pois.2D <- fit.ns(points = points, lims = rbind(c(0, 1), c(0, 1)),R = 0.5, dispersion="uniform",child.dist = child.dist)
```

# CRC data


# References