---
title: "Supplementary material: Estimating species distribution in highly dynamic populations using point process models"
author: ""
date: "February 2018"
output: md_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,tidy = TRUE, message = FALSE,tidy.opts=list(blank=FALSE, width.cutoff=60))
```

This supplementary material provides the code used to fit the spatio-temporal marked log-Gaussian Cox process discussed in *Estimating species distribution in highly dynamic populations using point process models* submitted to *Ecography*. There is an onlie version of this tutorial also avaiable [here](https://github.com/cmjt/examples/blob/master/Ecography_ESD.md).

Please note that the data cannot be supplied along with the supplementary material due to the protection status of the species. However, if readers do wish to access the data please contact [Andrea Soriano Redondo](A.Soriano-Redondo@exeter.ac.uk).

The \texttt{wetland} refered to below is a data frame of 65676 rows and 20 columns. Each row corresponds to an observation, and the columns picked out for the analysis correspond to the latitude and longitude of wetland locations, sum of breeding crane pairs on each wetland, wetland density, wetland area, a wetland density buffer and urban density buffer.


```{r packages,mess}
## required packages
library(INLA)

## load data. 
wetland <- read.csv("Wetland.csv")

```


Code section below extracts wetland coordinates and creates time index and binomial mark vector.

```{r data,fig.height=7}
## get coordinates (latitude and longitude of wetland locations)
locs <- cbind(wetland[,14],wetland[,15])
plot(locs,main = "Wetland locations in England and Wales", axes = FALSE, xlab  = "", ylab = "")

## transform time from 2000-2015 to 1-16
t.index <- (wetland$Year-min(wetland$Year))+1

## Crane presence as binomial mark (0 absent/ 1 present; sum of breeding crane pairs on each wetland)
mark <-ifelse( wetland$SUM_PAIR == 0, 0, 1)
mark.family <- "binomial"
```

This section sets up the covariate data frame as required for model fitting 

```{r covariates,eval = FALSE}
## Setting covariates

wetland$PA_ratio<-wetland$Perimeter/wetland$Area
## wetland density, wetland area, a wetland density buffer and urban density buffer.
cov1 <- wetland[,c(4,12,20,21)] 
cov1$Area_sc <- scale(cov1$Area)
cov1$PA_ratio_sc <- scale(cov1$PA_ratio)
cov1$Wet_density_nosea_sc <- scale(cov1$Wet_density_buf_NoSea)
cov1$Urb_density_nosea_sc <- scale(cov1$Urb_density_buf_NoSea)

## covariate data frame
covariates <- data.matrix(cov1[,c(5,6,7,8)])
# function that takes in a named  matrix of covariates with ncol equal to the number of covariates
# returns a list containing the effects ready to be read by (inla.stack) and a covariate formula (ready to be read
# by inla)
make.covs <- function(covariates){
  n.covs <- ncol(covariates)
  for(i in 1:n.covs){
    assign(colnames(covariates)[i],covariates[,i],envir = .GlobalEnv)
  }
  cov.effects <- sapply(colnames(covariates),get,simplify = FALSE)
  cov.form <- paste(colnames(covariates),collapse = " + ")
  return(list(cov.effects,cov.form))
}

m <- make.covs(covariates)
cov.effects <- m[[1]]
cov.form <- m[[2]]
```
The mesh is a requirement of the SPDE model fot the latent fields.

```{r mesh,eval = FALSE}
## make mesh for SPDE model
mesh <- inla.mesh.2d(loc = locs, cutoff = 0.6, max.edge = c(0.2,2))
```
This section sets priors used for hyperparameters of the model and sets up latent structures.

```{r priors etc,eval = FALSE}

## prior distributions
## prior for ar1 parameter
prior.rho = list(theta = list(prior='pccor1', param = c(0, 0.9))) 
## prior for interaction parameter
hyper = list(theta=list(prior='normal', param=c(0,10))) 

## Define SPDE model for latent fields
spde <-inla.spde2.matern(mesh = mesh, alpha = 2)
## number of observations
n <- nrow(locs)
## number of mesh nodes
nv <- mesh$n

temp <- t.index ## temporal dimentson
k <- (mesh.t <- inla.mesh.1d(temp))$n ## number of groups
## the response for the point pattern locations
y.pp <- rep(0:1, c(k * nv, n))
## create projection matrix for loacations
Ast <- inla.spde.make.A(mesh = mesh, loc = locs, group = temp, n.group = k)
## effect for LGCP used for point pattern
st.volume <- diag(kronecker(Diagonal(n = k),spde$param.inla$M0))
expected <- c(st.volume, rep(0, n))
## create field indecies
field.pp <- inla.spde.make.index('field.pp', n.spde = spde$n.spde, group = temp, n.group = k)
field.mark <- inla.spde.make.index('field.mark', n.spde = spde$n.spde, group = temp, n.group = k)
copy.field <- inla.spde.make.index('copy.field', n.spde = spde$n.spde, group = temp, n.group = k)
## temporal model "ar1"
ctr.g <- list(model='ar1',param = prior.rho) 

```

Here the data stacks, a requirement of \texttt{inla}, are constructed. They contain the information (i.e., covariates and fields) for each component of the joint response model.

```{r stack prep,eval = FALSE}
## Prepare data stacks for use in call to inla
## point process stack
stk.pp <- inla.stack(data=list(y=cbind(y.pp,NA), e=expected),
                     A=list(rBind(Diagonal(n=k*nv), Ast)),
                     effects=list(field.pp = field.pp))
## mark stack
stk.mark <- inla.stack(data=list(y=cbind(NA,mark)),
                       A=list(Ast, Ast,1),
                       effects=list(field.mark = field.mark, 
                                    copy.field = copy.field, cov.effects = cov.effects))
## combine data stacks
stack <- inla.stack(stk.pp,stk.mark)

```
Finally the formula is defined and \texttt{inla} is called to fit the joibt spatio-temporalmarked log-Gaussian Coz process model.

```{r inla,eval = FALSE}

## formula for model as given by equation X in article

formula <- y ~ 0 + Area_sc + PA_ratio_sc + Wet_density_nosea_sc + Urb_density_nosea_sc +
  f(field.pp, model=spde, group = field.pp.group, control.group=ctr.g) +
  f(field.mark, model=spde, group = field.mark.group , control.group=ctr.g) +
  f(copy.field, copy = "field.pp", fixed=FALSE, hyper = hyper )

fit <- inla(formula, family = c("poisson",mark.family),data=inla.stack.data(stack),
           E=inla.stack.data(stack)$e,
           control.predictor=list(A=inla.stack.A(stack)),
           control.inla=list(strategy='gaussian',int.strategy = 'eb'),
           verbose = TRUE) ## verbose = TRUE will print out modelling output to console
```