---
title: "shared"
author: "cmjt"
date: "22 November 2017"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = FALSE)
library(devtools)
load_all("~/Git/lgcpSPDE")

```

## Applications of marked LGCPs

All funtions used below are available in the `lgcpSPDE` package [here](https://github.com/cmjt/lgcpSPDE), which can be installed by using `devtools::install_github("cmjt/lgcpSPDE")`.

### Spatial dynamics of terrorism activity worldwide

The terrorism data is available within the `lgcpSPDE` package and can be found by running `data(terrorism)`. In addition a `SpatialPolygonsDataFrame` of the world used in the construction of the mesh can be found by running `data(world)`.

Terrorism activity, 2010--2016, is overlaid onto a map of the world below.

```{r terrorism data},eval = TRUE,echo = FALSE}
data(terrorism)
data(world)
plot(world)
points(terrorism $longitude, terrorism$latitude, pch = 20)

```

```{r terrorism prep,echo = TRUE, eval = TRUE}
head(terrorism)
locs <- as.matrix(terrorism[,5:7])
t.index <- (terrorism$iyear - min(terrorism$iyear)) + 1
mark <- terrorism$fatal ## 1 = fatalities, 2 = no fatalities
```

```{r terrorism model fit, eval = FALSE, echo = TRUE}
fit <- fit.marked.lgcp(mesh = mesh, locs = locs,t.index = t.index, 
                   mark = mark, mark.family = "binomial",
                   verbose=TRUE, 
                   hyper = list(theta = list(prior = "normal", param = c(0, 10))),
                   link = 2) ## use quick strategy first

T.fit.imp <- fit.marked.lgcp(mesh = mesh, locs = locs, t.index = t.index, 
                             mark = mark, mark.family = "binomial",
                             verbose=TRUE, 
                             hyper = list(theta = list(prior = "normal", param = c(0, 10))),
                             control.mode = list(result = fit,restart = TRUE),
                             control.inla = list(strategy='laplace'),
                             link = 2)## improve using starting values from first model
```