---
title: "Supplementary material: Estimating species distribution in highly dynamic populations using point process models"
author: ""
date: ""
output: md_document
header-includes:
- \usepackage{natbib}
- \usepackage{amsmath}
bibliography: bib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,tidy = TRUE, message = FALSE,
                      tidy.opts=list(blank=FALSE, width.cutoff=60),warning = FALSE)

```

This supplementary material provides the code used to fit the spatio-temporal marked log-Gaussian Cox process discussed in *Estimating species distribution in highly dynamic populations using point process models* submitted to *Ecography*. There is an online version of this tutorial also available [here](https://github.com/cmjt/examples/blob/master/species_distribution.md).

Please note that the data used in the article cannot be supplied along with the supplementary material due to the protection status of the species (if readers do wish to access the data please contact [Andrea Soriano Redondo](A.Soriano-Redondo@exeter.ac.uk)). However, we supply data simulated from the model to illustrate the methodology detailed in the article. Alongside this we supply functionality so that the readers can fit the models discussed; the procedure for this is outlined below.

The functionality is provided in the suite of R functions, **lgcpSPDE**, available [here](https://github.com/cmjt/lgcpSPDE); to install the **lgcpSPDE** package from GitHub run the following R code.
```{r, eval = FALSE}
devtools::install_github("cmjt/lgcpSPDE") 
```

## The data

The simulated example data used here, contained in the R object **cranes**, is included in the **lgcpSPDE** package. This dataframe has 5052 wetland observations and 10 variables:

- **Wetland_Identity**, a numeric wetland ID;
- **Lon**, longitude epicentre location of wetland;
- **Lat**, latitude epicentre location of wetland;
- **Area**, Area of wetland in $m^2$;
- **Perimiter**, Wetland perimiter in $m$;
- **Wet_density_buf_NoSea**, Surrounding wetland density;
- **Urb_density_buf_NoSea**, Surrouning urban density;
- **Year**, Year of observation (i.e., 2014 or 2015);
- **mark**, A simulated binary mark indicating presence of a Grus Grus breeding pair at the wetland;
- **PA_ratio**, Wetland perimiter to area ratio.

To load and inspect the data run the R following code.

```{r data}
library(lgcpSPDE)
data(cranes)
str(cranes) ## to see further details run ?cranes
```

The point pattern (of wetlands) and assocciated mark (binary presence or absence of a pair of breeding cranes) is shown below.

```{r plot, echo = FALSE,fig.width = 10,fig.cap = "Epicentre locations of wetlands in England (i.e., potentially suitable habitats for Grus Grus). Black plotting characters indicate the wetlands that were (simulated to be) occupied by a breeding pair in 2014 an/or 2015."}
library(ggplot2)
UK <- map_data(map = "world", region = "UK")
UK <- subset(UK, UK$subregion == "Great Britain")
plot(UK[,1:2],type = "l",axes = FALSE,xlab = "", ylab = "",asp = 1)
col <- ifelse(cranes$mark == 1, "black","grey")
points(cranes$Lon,cranes$Lat,pch = 20,col = col,cex = 0.5)
legend("top",bty = "n",pch = 20, col = c("black","grey"),legend = c("occupied","un-occupied"),cex = 0.5)
```

## Modelling

In order to fit a marked spatio-temporal model to this example data, as detailed in the article, first construct the required R objects as below.

```{r setup data}
locs <- as.matrix(cranes[,2:3]) ## matrix of wetland epicentre locations
mark <- cranes$mark ## crane breeding pair presence as binary mark (0 absent/ 1 present)
table(mark,cranes$Year) ## 10 occupied sites in 2014, 20 in 2015
mark.family <- "binomial" 
t.index <- (cranes$Year-min(cranes$Year)) + 1 ## create year index
## Create covariate data frame (scaled)
covariates <- data.frame(Area_sc = scale(cranes$Area), PA_ratio_sc = scale(cranes$PA_ratio),
                          Wet_density_nosea_sc = scale(cranes$Wet_density_buf_NoSea),
                          Urb_density_nosea_sc = scale(cranes$Urb_density_buf_NoSea))
## construct the mesh ( course just for illustration purposes)
mesh <- inla.mesh.2d(loc = locs, cutoff = 0.6, max.edge = c(0.2,2)) 
```


The triangulation (mesh) is an integral part of the INLA-SPDE methodology. The mesh provides the basis on which the approximation of the Gaussian random fields are constructed. Choices regarding the sparsity of the mesh as well as its shape are not simple to make. The mesh should (among other things)

- reflect the spatial structure you wish to capture, and

- provide a decent representation of the spatial domain.

Futher details on mesh construction can be found in @lindgren2011explicit.

This supplementary material is simply to illustrate how the models discussed in the article were fitted and to enable readers to go through the same procedure in a timely manner; therefore, the mesh we construct (shown below) is badly formed and too sparse---simply aid in computation time.


```{r mesh, fig.width = 10, echo = FALSE,fig.cap = "Example Delauney triangulation (mesh) required for the marked spatio-temporal log-Gaussian Cox process model. Overlain are the wetland locations, black occupies and grey un-occupied, and a spatial polygon of the UK (not including Northen Irelan)."}

plot(mesh,main = "", asp = 1)
lines(UK,lwd = 2)
points(cranes$Lon,cranes$Lat,pch = 20,col = col,cex = 0.5)


```

To fit the model the **fit.marked.lgcp()** from **lgcpSPDE** is called. This function offers a number of options, run 
```{r, eval = FALSE}
args(fit.marked.lccp)
```
in R to see all options.

To fit the model discussed in *Estimating species distribution in highly dynamic populations using point process models* to this simulated example data we use the following arguments,

- **mesh**, the Delauney trangulation discussed above;
- **locs**, a 2 times n matrix of wetland locations;
- **t.index**, a vector of year indecies for the n observations;
- **covariates**, a data frame of the named covariates;
- **mark**, a vector of the binary mark values;
- **mark.family**, the distribution that the marks are assumed to be a realisation of (i.e., "binomial");
- **prior.range** \& **prior.sigma**, peanalised complexity priors on the spatial range and standard deviation (see Appendix A2 for more details);
- **pp.int**, logical indicating if the linear predictor for the point pattern component of the model should contain an intercept term. If TRUE the parameter $\alpha_0$ is estimated;
- **mark.int**, logical indicating if the linear predictor for the mark component of the model should contain an intercept term. If TRUE the parameter $\beta_0$ is estimated.

The following finction call in R will fit a marked spatio-temporal log-Gaussian Cox process model, as detailed in the article, to the example data.

```{r fitting function,eval = TRUE}
fit <- fit.marked.lgcp(mesh = mesh, locs = locs, t.index = t.index, 
                       covariates = covariates, mark = mark,
                       mark.family = mark.family,
                       prior.range = c(4,0.5),
                       prior.sigma = c(1,0.05),
                       pp.int = TRUE, mark.int = TRUE)
```

### Model inference

Use the **summary** utility function to view the posterior estimates of the model parameters. The fixed effect parameters relate to the coefficients of the fixed effects. In the model fitted here these parameters are 

- $\alpha_0$, the intercept for the point pattern component of the model;
- $\beta_0$, the intercept for the mark component of the model;
- the coefficients of the named covariates of the mark component.

The parameters (hyper-parameters) of the latent fields are the spatial range and standard deviation of the named field along with the $\rho$ parameter for the assumed AR(1) temporal process. The fields in this model are **field.pp**, $\mathbf{z}(\mathbf{s},t)$ in the article (i.e., reflects the spatially varying intensity of wetlands), and **field.mark**, $\mathbf{g}(\mathbf{s},t)$ in the artile (i.e., conditional on the spatial intensity of the wetlands this reflects the spatially varying process of presence of a breeding crane pairs). As the Gaussian field $\mathbf{z}(\mathbf{s},t)$ is shared between both components of the model the interaction parameter $\beta$ (named *Beta for copy.field*) is estimated.

```{r summary}
summary(fit)$fixed[,1:5] ## print out some summary information for fixed effects
summary(fit)$hyperpar[,1:5] ## some summary information for parameters of the latent fields
## extract the random fields from the model object over the n.t = 2 years 
## This returns a list of matrices
fields <- find.fields(fit, mesh, n.t = 2) 
```

The **fields** object above is a named list of matrices of each Gaussian field for each year. Below $\mathbf{z}(\mathbf{s},t)$ (left) and $\mathbf{g}(\mathbf{s},t)$ (right) are plotted for the first year on the link scale.

```{r plots, echo = FALSE,fig.width = 8, fig.cap = "Posterior means of the random fields on the link scales. Left plot shows the random field of the point process intensity. Right plot shows the mark specific random field once the spatial structure of the wetlands, and fixed effects, have been accounted for."}
par(mfrow = c(1,2))
image.plot(fields[[1]][[1]], axes = FALSE)
image.plot(fields[[2]][[1]], axes = FALSE)
```

\newpage
## References